---
title: "R Session 1 Practice Sheet"
author: "CDN team"
date: "6/4/2024"
output: html_document
---

Hello and welcome to your first R homework assignment! You should reference both
your notes from the session and the [Base-R cheatsheet](https://github.com/rstudio/cheatsheets/blob/main/base-r.pdf) 
to answer the following questions.

## Part 1: Objects + Basic Math
*Remember to reference the [cheat sheet](https://github.com/rstudio/cheatsheets/blob/main/base-r.pdf)
 if you get stuck*  
 
1. Generate a sequence from 3 to 71 and name it `my_sequence.`  
```{r}

```

2. Generate a sequence from 3 to 71 that only has odd numbers, and call it `odds`.  
```{r}

```

3. Find the sum of `odds`  
```{r}

```

4. Find the product of the 7th and 11th element of `odds`  
```{r}

```

## Part 2: `for` Loops and `if` statements
*Remember to reference the [cheat sheet](https://github.com/rstudio/cheatsheets/blob/main/base-r.pdf)
if you get stuck*  
 
5. Write a for loop that prints out every fifth element of `odds` (i.e., the 
5th element, the 10th element, the 15th element, etc)
```{r}

```

6. Using a for loop, construct a vector of length 10, where the first element 
is `100`, and every following element is double the square-root of the previous 
one.
```{r}
n <- 100
vec <- c()

for (i in 1:10) {
    
    if (i == 1) {
        vec <- c(vec, n)
    } else {
        vec <- c(vec, 2 * sqrt(vec[i - 1]))
    }
   
}
vec
```

7. Write an `if / else` statement that prints "Yay" if some value `x` is 
greater than 10 and "Awesome" otherwise. Try setting `x` to 
different values and try running it.
```r
x <- 5

if () {
    
}
```

## Part 3: Functions
*Remember to reference the [cheat sheet](https://github.com/rstudio/cheatsheets/blob/main/base-r.pdf)
 if you get stuck*  
 
Functions are defined like:
```r
name_of_function <- function(parameter) {
  # do_something...
}
```

Here's an example:
```r
foo <- function(x) {
  x * 2 + 3 / 4
}
```

The last calculation that happens in the function will be returned by default. 
If you want, you can add an explicit `return` statement, like this:

```r
foo <- function(x) {
  return(x * 2 + 3 / 4)
}
```
or
```r
foo <- function(x) {
  y <- x * 2 + 3 / 4
  return(y)
}
```
But you don't need to - all three of those examples will work exactly the same.

8. Try it out! Define the above `foo` function in your environment, and try 
running it with different parameters.
```{r}

```

9. Add a second parameter (`y`) to `foo` that raises `x` to the `yth` power.
```{r}

```

10. Moving on from `foo`, define a new function `foo2`, that returns `TRUE` if 
the input is greater than 5.
```{r}

```

To set a default value in a function, define it like:
```r
foo <- function(x = 1) {
  x * 2 + 3 / 4
}
```

11. Define and run the above `foo` function in your environment, and try running
it without any parameters `foo()` and with other parameters `foo(3)`.
```{r}

```

12. Edit your `foo2` function definition to have a default parameter value of 3.
```{r}

```

## THAT'S ALL!

We'll send around an answer key in a few days, but take a stab at getting through
 what you can on your own. And if you have time, read Hadley Wickham's paper 
 on [tidy data](https://cran.r-project.org/web/packages/tidyr/vignettes/tidy-data.html)!